The Constant-Q Transform Spectral Envelope Coefficients: A Timbre Feature Designed for Music

### 2021-06-23 IEEE SP Magazine/06-2021 - Draft/article.tex
Almost done with the extraction subsection.

This convolution process can be thought of a source-filter model \cite{fant1970} which is here not applied in the time domain but in the frequency domain, with the source and the filter being the pitch component and the envelope, respectively.

@@

Assuming that pitch and timbre are independent, this implies that the same musical object at different pitches would have a similar envelope but a shifted pitch component (while two different musical objects at the same pitch would have different envelopes but a similar pitch component). This is summarized in Equation \ref{eq:observation1}, where $X$, $E$, $P$ and $X'$, $E'$, $P'$ represent the log-spectrum, envelope, and pitch component for a musical object and for a pitch-shifted version of the same musical object, respectively.
\begin{equation}
\label{eq:observation1}
\begin{split}
\begin{cases}
X = E * P \\
X' = E' * P' \\
\end{cases}
\Rightarrow E \approx E'
\end{split}
\end{equation}

@@

This implies that the FT of the log-spectrum is equal to the point-wise product of the FT of the envelope and the FT of the pitch component. Given the first observation, this further implies that the FT of the envelope for a musical object and for a pitch-shifted version of it would be equal. This is summarized in Equation \ref{eq:observation2}, where $\mathcal{F}(.)$ represents the FT function.
\begin{equation}
\label{eq:observation2}
\begin{split}
\begin{cases}
\mathcal{F}(X) = \mathcal{F}(E) \cdot \mathcal{F}(P) \\
\mathcal{F}(X') = \mathcal{F}(E') \cdot \mathcal{F}(P') \\
\end{cases}
\Rightarrow \mathcal{F}(E) \approx \mathcal{F}(E')
\end{split}
\end{equation}

@@

This implies that the magnitude of the FT of the log-spectrum for a musical object and for a pitch-shifted version of it would be equal. This is summarized in Equation \ref{eq:observation3}, where $|.|$ and $Arg(.)$ represent the modulus and argument, respectively, for a complex array. 
\begin{equation}
\label{eq:observation3}
\begin{split}
\begin{cases}
\mathcal{F}(X) = |\mathcal{F}(X)| \cdot e^{j Arg(\mathcal{F}(X))} \\
\mathcal{F}(X') = |\mathcal{F}(X')| \cdot e^{j Arg(\mathcal{F}(X'))} \\
\end{cases}
\Rightarrow |\mathcal{F}(X)| \approx |\mathcal{F}(X')|
\end{split}
\end{equation}

Given the previous observations, we can therefore conclude that the FT of the envelope could be approximated by the magnitude of the FT of the log-spectrum, while the FT of the pitch component could be approximated by the phase component. This finally gives us the estimates for the envelope and the pitch component, after taking their inverse FTs, as shown in Equation \ref{eq:conclusion}, where $\mathcal{F}^{-1}(.)$ represents the inverse FT function.
\begin{equation}
\label{eq:conclusion}
\begin{split}
\Rightarrow 
\begin{cases}
\mathcal{F}(E) \approx |\mathcal{F}(X)| &\Rightarrow E \approx \mathcal{F}^{-1}(|\mathcal{F}(X)|) \\
\mathcal{F}(P) \approx e^{j Arg(\mathcal{F}(X))} &\Rightarrow P \approx \mathcal{F}^{-1}(e^{j Arg(\mathcal{F}(X))}) \\
\end{cases}
\end{split}
\end{equation}

Figure \ref{fig:deconvolution} shows and example of deconvolution of the (power) CQT-spectrogram into its envelope and pitch component. The CQT-spectogram was computed from a 48-second audio signal created by concatenating 12 4-second acoustic bass notes, playing successively from C1 (32.70 Hz) to B1 (61.74 Hz). The notes come from the NSynth dataset \cite{engel2017}, for instrument \texttt{bass\_acoustic\_000}, MIDI numbers \texttt{024} to \texttt{035}, and velocity number \texttt{075}. As we can see, the envelope gets pitch-normalized, shifting all the frequency components to the origin of the log-spectrum (essentially bringing all the notes to C1), while the pitch component gets energy-normalized, leaving mostly the fundamental frequency of the notes.

\begin{figure*}[htp]
    \centering
    \includegraphics[width=\textwidth]{deconvolution.png}
    \caption{Deconvolution of the (power) CQT-spectrogram of 12 acoustic bass notes playing from C1 to B1, into a pitch-independent envelope and an energy-normalized pitch component.}
    \label{fig:deconvolution}
\end{figure*}

This deconvolution process can also be thought of the normalization of the log-spectrum by the magnitude of its FT (which here would correspond to the FT of the envelope) leading to a sharper log-spectrum (which here would correspond to the pitch component), in the manner of the generalized cross-correlation phase transform (GCC-PHAT) method which aims at normalizing a cross-correlation function by its magnitude spectrum to sharpen the cross-correlation peaks \cite{knapp1976}.

@@

%HERE!!!
Once the CQT-spectrum has been decomposed into a spectral envelope and a pitch component, we can then assumed that most of the pitch information has been removed from the envelope which now mostly contains the timbre information. We can think of the spectral envelope as a pitch-normalized log-spectrum where the frequency components have been shifted down to the origin of the spectrum.

Equation \ref{eq:extraction} $O_r$ the octave resolution, or the number of frequency channels per octave (typically, 12, 24, 36, etc.), $N_c$ the number of coefficients

\begin{equation}
\label{eq:extraction}
i = O_r \log_2(k) \text{ for } 1 \le k \le N_c
\end{equation}

>>>>

This convolution process can be thought of as a source-filter model \cite{fant1970} which is here not applied in the time domain but in the frequency domain, with the source and the filter being the pitch component and the envelope, respectively.

@@

Assuming that pitch and timbre are independent, this implies that the same musical object at different pitches would have a similar envelope but a shifted pitch component (while two different musical objects at the same pitch would have different envelopes but a similar pitch component). This is summarized in Equation \ref{eq:observation1}, where $X$, $E$, $P$ and $X'$, $E'$, $P'$ represent the log-spectrum, envelope, and pitch component for a musical object and for a pitch-shifted version of the same musical object, respectively.
\begin{equation}
\label{eq:observation1}
\begin{split}
& \begin{cases}
X = E * P \\
X' = E' * P' \\
\end{cases} \\
& \Rightarrow E \approx E'
\end{split}
\end{equation}

@@

This implies that the FT of the log-spectrum is equal to the point-wise product of the FT of the envelope and the FT of the pitch component. Given the first observation, this further implies that the FT of the envelope for a musical object and for a pitch-shifted version of it would be equal. This is summarized in Equation \ref{eq:observation2}, where $\mathcal{F}(.)$ represents the FT function.
\begin{equation}
\label{eq:observation2}
\begin{split}
& \begin{cases}
\mathcal{F}(X) = \mathcal{F}(E) \cdot \mathcal{F}(P) \\
\mathcal{F}(X') = \mathcal{F}(E') \cdot \mathcal{F}(P') \\
\end{cases} \\
& \Rightarrow \mathcal{F}(E) \approx \mathcal{F}(E')
\end{split}
\end{equation}

@@

This implies that the magnitude of the FT of the log-spectrum for a musical object and for a pitch-shifted version of it would be equal. This is summarized in Equation \ref{eq:observation3}, where $|.|$ and $Arg(.)$ represent the modulus and argument, respectively, for a complex array. 
\begin{equation}
\label{eq:observation3}
\begin{split}
& \begin{cases}
\mathcal{F}(X) = |\mathcal{F}(X)| \cdot e^{j Arg(\mathcal{F}(X))} \\
\mathcal{F}(X') = |\mathcal{F}(X')| \cdot e^{j Arg(\mathcal{F}(X'))} \\
\end{cases} \\
& \Rightarrow |\mathcal{F}(X)| \approx |\mathcal{F}(X')|
\end{split}
\end{equation}

Given the previous observations, we can therefore conclude that the FT of the envelope could be approximated by the magnitude of the FT of the log-spectrum, while the FT of the pitch component could be approximated by the phase component. This finally gives us the estimates for the envelope and the pitch component, after taking their inverse FTs, as shown in Equation \ref{eq:conclusion}, where $\mathcal{F}^{-1}(.)$ represents the inverse FT function.
\begin{equation}
\label{eq:conclusion}
\begin{split}
& \Rightarrow 
\begin{cases}
\mathcal{F}(E) \approx |\mathcal{F}(X)| \\
\mathcal{F}(P) \approx e^{j Arg(\mathcal{F}(X))} \\
\end{cases} \\
& \Rightarrow
\begin{cases}
E \approx \mathcal{F}^{-1}(|\mathcal{F}(X)|) \\
P \approx \mathcal{F}^{-1}(e^{j Arg(\mathcal{F}(X))}) \\
\end{cases}
\end{split}
\end{equation}

Figure \ref{fig:deconvolution} shows an example of deconvolution of a (power) CQT-spectrogram into its envelope and pitch component. The CQT-spectogram was computed from a 48-second audio signal created by concatenating 12 4-second notes from an acoustic bass playing from C1 (32.70 Hz) to B1 (61.74 Hz) in ascending order. The notes come from the NSynth dataset \cite{engel2017} and correspond to instrument id \texttt{bass\_acoustic\_000}, MIDI numbers \texttt{024} to \texttt{035}, and velocity number \texttt{075}. As we can see, the envelope looks as if the CQT-spectrogram has been normalized in pitch, with all the notes being brought down to the lowest frequency, here C1, while the pitch component looks as if the CQT-spectrogram has been stripped down from all its energy, leaving mostly the fundamental frequencies of the notes.

\begin{figure*}[htp]
    \centering
    \includegraphics[width=\textwidth]{deconvolution.png}
    \caption{Deconvolution of the CQT-spectrogram (left plot, shown in dB) of 12 acoustic bass notes playing from C1 to B1, into a pitch-independent envelope (middle plot, shown in dB) and an energy-normalized pitch component (right plot, shown in $[0, 1]$).}
    \label{fig:deconvolution}
\end{figure*}

This deconvolution process can also be thought of as the normalization of the log-spectrum by the magnitude of its FT (which here would correspond to the FT of the envelope) leading to a sharper log-spectrum (which here would correspond to the pitch component), in the manner of the generalized cross-correlation phase transform (GCC-PHAT) method which aims at normalizing a cross-correlation function by its magnitude spectrum to sharpen the cross-correlation peaks \cite{knapp1976}.

@@

Once the CQT-spectrum has been decomposed into an envelope and a pitch component, we can then assume that most of the pitch information has been removed from the envelope which now mostly contains the timbre information. As shown in the middle plot of Figure \ref{fig:deconvolution}, the envelope can be thought of as a pitch-normalized log-spectrum where the spectral components have been shifted down to the origin of the spectrum. If we know what octave resolution was used when computing the CQT-spectrum, i.e., the number of bins per octave, we can easily infer the locations of the harmonics in the envelope, which should then judiciously characterize the timbre of the musical instrument in question. Hence, the CQT-SECs. Equation \ref{eq:extraction} shows how to derive the indices of the harmonics given $O_r$, the octave resolution, and $N_c$, the number of desired coefficients, and consequently, the CQT-SECs from the envelope $E$. 
\begin{equation}
\label{eq:extraction}
\begin{split}
\begin{cases}
i = O_r \log_2(k) \\
\text{CQT-SEC}_k = E(i) \\
\end{cases}
\quad 1 \le k \le N_c
\end{split}
\end{equation}
